---
title: "README"
author: "Jens Riis Baalkilde"
date: "December 11, 2024"
output: 
  md_document:
      variant: gfm
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


bmd
========================================================

This is an R package for benchmark dose (BMD) estimation, which expands upon the functionality of the drc package. 

This package is currently maintained by Jens Riis Baalkilde and Signe Marie Jensen, Department of Plant and Environmental Sciences, University of Copenhagen.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(drcData)
library(drc)
library(bmd)
```


Installation
-------------------------

Install the bmd package from GitHub. The bmd package works best with the drc package installed from GitHub as well.

```{r installation, eval=FALSE}
install.packages("devtools")
devtools::install_github("DoseResponse/drcData")
devtools::install_github("DoseResponse/drc")
devtools::install_github("DoseResponse/bmd")
```

Key features
-------------------------

The bmd package is a natural extension of the drc package. Key features of the bmd package includes:

- Benchmark dose (BMD) estimation for quantal, continuous, count and ordinal data for a wide range of definitions of BMD.
  - A range of confidence intervals for BMD estimates, including Wald, Profile and Bootstrap intervals.
  - BMD estimation based on model averaging.
  - BMD estimation for multiple dose-response curves.
- Utility functions for plotting dose-response curves and BMD estimates.

Example usage
-------------------------

In the following it is demonstrated how to use the functions in the bmd package on two data sets with a continuous response variable and one data set with a quantal response.

### Example 1 (secalonic)

First, the data set from the drcData package is loaded, and a dose-response model is fitted. 

```{r secalonic}
data("secalonic")
secalonic.LL.4 <- drm(rootl ~ dose, data = secalonic, fct = LL.4())
```

Users familiar with the drc package already know how straight-forward it is to plot dose-response models using the plot() function. The qplotDrc() function in the bmd package mimics the same functionality based on ggplot2. 

```{r secalonic_plot}
plot(secalonic.LL.4, main = "Secalonic model plotted by basic R plotting")

qplotDrc(secalonic.LL.4) + ggplot2::labs(title = "Secalonic model plotted by ggplot2")
```

Similar to the old method of plotting dose-response curves, qplotDrc() features several options for customisation.
```{r secalonic_plot_features}
qplotDrc(secalonic.LL.4, type = "all")
qplotDrc(secalonic.LL.4, type = "obs")
qplotDrc(secalonic.LL.4, type = "bars")
qplotDrc(secalonic.LL.4, type = "confidence")
```


#### Example 1: BMD estimation

A wide range of BMD definitions are implemented in the bmd package. In the following, the BMD and BMDL based on the "relative risk" definition recommended by EFSA, with a BMR = 10% are computed.

```{r secalonic_bmd}
bmd(secalonic.LL.4, bmr = 0.1, backgType = "modelBased", def = "relative")
```

The default interval is a Wald type confidence interval. If a profile likelihood confidence interval is preferred, this can be specified by the "interval" argument.

```{r secalonic_bmd_profile}
bmd(secalonic.LL.4, bmr = 0.1, backgType = "modelBased", def = "relative", interval = "profile")
```


The current state-of-the-art definition of the BMD is the hybrid definition. In the following, the BMD and BMDL based on the "excess risk" definition based on the hybrid method with BMR = 10%, and adverse background level set to the estimated background level minus 1SD is calculated.

```{r secalonic_bmd_hybrid}
bmd(secalonic.LL.4, bmr = 0.1, backgType = "hybridSD", backg = 1, def = "hybridExc")
```

For technical reasons, profile intervals for BMD estimates based on the hybrid method, are not currently available. 

#### Example 1: BMD plotting

The qplotBmd() function ca be used to plot BMD along with the dose-response curve.

```{r secalonic_bmd_plot}
qplotBmd(bmd(secalonic.LL.4, bmr = 0.1, backgType = "modelBased", def = "relative", display = FALSE)) # display = FALSE hides output from bmd function
```


#### Example 1: BMD estimation using MA

An additional set of dose-response models are fitted.

```{r}
secalonic.LL.3 <- drm(rootl ~ dose, data = secalonic, fct = LL.3())
secalonic.LN.3 <- drm(rootl ~ dose, data = secalonic, fct = LN.3())
secalonic.LN.4 <- drm(rootl ~ dose, data = secalonic, fct = LN.4())
secalonic.W1.3 <- drm(rootl ~ dose, data = secalonic, fct = W1.3())
secalonic.W1.4 <- drm(rootl ~ dose, data = secalonic, fct = W1.4())
secalonic.W2.3 <- drm(rootl ~ dose, data = secalonic, fct = W2.3())
secalonic.W2.4 <- drm(rootl ~ dose, data = secalonic, fct = W2.4())

secalonic.modelList <- list(secalonic.LL.3, secalonic.LL.4, secalonic.LN.3, secalonic.LN.4,
                            secalonic.W1.3, secalonic.W1.4, secalonic.W2.3, secalonic.W2.4)
```

BMD can now be estimated by MA by using the function bmdMA. Type of model weights need to be specified as well as MA type. 

```{r secalonic_bmdMA, warning=FALSE}
bmdMA(secalonic.modelList, modelWeights = "AIC", type = "Buckland", bmr = 0.1, backgType = "modelBased", def = "relative")

set.seed(123)
bmdMA(secalonic.modelList, modelWeights = "AIC", type = "bootstrap", bmr = 0.1, backgType = "modelBased", def = "relative", R = 500, progressInfo = FALSE)
bmdMA(secalonic.modelList, modelWeights = "AIC", type = "Buckland", bmr = 0.1, backgType = "hybridSD", def = "hybridExc", backg = 1)
```




### Example 2 (S.alba.comp, two dose-response curves)

An experiment was conducted where some pots were treated with the herbicide Bentazone, and other pots were treated with Glyphosate. A model with separate curves for each herbicide is fitted as follows.

```{r S.alba.comp, warning=FALSE}
data("S.alba.comp")
S.alba.comp.LL.4 <- drm(drymatter ~ dose, curveid = herbicide, data = S.alba.comp, fct = LL.4())
```

To plot the function, qplotDrc can be used:
```{r S.alba.comp_plot}
qplotDrc(S.alba.comp.LL.4)
qplotDrc(S.alba.comp.LL.4, col = TRUE, type = "confidence") +
  qplotDrc(S.alba.comp.LL.4, col = TRUE, type = "obs", add = TRUE)$obsLayer
```

This also illustrates how qplotDrc() can be layered to produce custom plots, by using the argument "add = TRUE", and choosing either the layer of observations ("obsLayer"), the layer with the dose-response curve ("curveLayer"), or the layer with the confidence band around the curves ("confBandLayer").

#### Example 2: BMD estimation.

BMD and BMDL can be estimated for the individual curves:

```{r S.alba.comp_bmd}
bmd(S.alba.comp.LL.4, bmr = 0.1, backgType = "modelBased", def = "relative")
bmd(S.alba.comp.LL.4, bmr = 0.1, backgType = "hybridSD", backg = 1, def = "hybridExc")
```


### Example 3 (acute.inh, binomial response)

20 animals were exposed to 4 doses of an unknown substance, 5 animals per dose. The number of dead animals for each dose were recorded. In the following, the data is loaded, and a 2-parameter is fitted to the data. Subsequently, a plot of the resulting dose-response curve is created using qplotDrc().

```{r acute.inh, warning=FALSE}
data("acute.inh")
acute.inh.LL.2 <- drm(num.dead/total ~ dose, weights = total, data = acute.inh, fct = LL.2(), type = "binomial")

qplotDrc(acute.inh.LL.2)
```




#### Example 3: BMD Estimation

For binomial response data, the "excess" and "additional" BMD definitions are available. When the background response is $0$, they are identical:

```{r acute.inh_bmd}
bmd(acute.inh.LL.2, bmr = 0.1, backgType = "modelBased", def = "additional")
bmd(acute.inh.LL.2, bmr = 0.1, backgType = "modelBased", def = "excess")
```



```{r acute.inh_bmd_plot}
qplotBmd(bmd(acute.inh.LL.2, bmr = 0.1, backgType = "modelBased", def = "additional", display = FALSE)) +
  ggplot2::scale_x_continuous(limits = c(0,2100), trans = scales::transform_pseudo_log(2000))
```




