---
title: "bmd Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bmd_Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(drcData)
library(drc)
library(bmd)
```


## Package Summary

This R package focuses on Benchmark Dose (BMD) modeling, which is commonly used in toxicology and risk assessment to determine the dose of a substance that corresponds to a specific level of response or risk. The package provides a  comprehensive range of functionalities. It supports various model types, confidence interval methods, and model averaging approaches.

### Core Functionalities

1. **Dose-Response Modeling**:
   - Fits various dose-response models including logistic, log-normal, Weibull, and fractional polynomial models
   - Supports continuous, binomial, Poisson, and negative binomial response types
   - Handles both single and multiple curves/groups

2. **BMD Estimation**:
   - Calculates benchmark doses (BMDs) and their lower confidence limits (BMDLs)
   - Implements multiple BMD definitions: excess, additional, relative, extra, added, hybrid excess, hybrid additional, and point
   - Provides various background types: model-based, absolute, hybridSD, and hybridPercentile

3. **Confidence Interval Methods**:
   - Delta method
   - Profile likelihood
   - Bootstrap (parametric, nonparametric)
   - Sandwich estimator
   - Inverse regression

4. **Model Averaging**:
   - Implements model averaging for BMD estimation via `bmdMA()`
   - Supports various weighting schemes: AIC, BIC, and stacking (cross-validation)
   - Implements different averaging approaches: Kang, Buckland, curve, and bootstrap

5. **Special Model Types**:
   - Ordinal response models (`drcOrdinal`)
   - Heterogeneous variance models (`drcHetVar`)
   - Mixed models with random effects (`drcMMRE`)

6. **Trend Testing**:
   - Implements monotonicity tests (Jonckheere-Terpstra, Bartholomew)
   - Provides trend tests (Williams, Shirley, Tukey)

7. **Visualization**:
   - Plot methods for BMD objects
   - ggplot2-based plotting functions (`qplotDrc`, `qplotBmd`)

## Usage of the Package

The package follows a workflow typically seen in dose-response analysis:

1. **Fit dose-response models** using functions like `drm()`, `drcOrdinal()`, `drcHetVar()`, or `drcMMRE()`
2. **Estimate BMDs** using `bmd()` or specialized functions like `bmdOrdinal()`, `bmdHetVar()`
3. **Apply model averaging** with `bmdMA()` if multiple models are considered
4. **Visualize results** using plot methods or specialized plotting functions
5. **Test for trends** using `trendTest()` or `monotonicityTest()`

### Example usage:

```{r}
# Fit dose-response model

data("secalonic")
model <- drm(rootl ~ dose, data = secalonic, fct = LL.4())
## model <- drm(response ~ dose, data = mydata, fct = LL.4())

# Calculate BMD
bmd_result <- bmd(model, bmr = 0.1, backgType = "modelBased", def = "relative", interval = "profile")

# Plot result
plot(bmd_result)
## note that ggplot2 style plot can be generated using qplot function
## qplotBmd(bmd(model, bmr = 0.1, backgType = "modelBased", def = "relative", display = FALSE)) # display = FALSE hides output from bmd function
```

#### Fit multiple models for model averaging

```{r}
# Fit multiple models for model averaging

# Apply model averaging (if multiple models are fitted)
model1 <- model
model2 <- drm(rootl ~ dose, data = secalonic, fct = LN.3())
model3 <- drm(rootl ~ dose, data = secalonic, fct = LN.4())
model_list <- list(model1, model2, model3)
bmd_ma <- bmdMA(model_list, modelWeights = "AIC", backgType = "modelBased", bmr = 0.1, 
                def =  "relative" , type = "bootstrap")
print(bmd_ma)
```

